Adapting the MVDR to a GPU for active sonar systems of different sizes.

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	02:39:33	(Wed23.Jan)

The beamspace MVDR formulation very closely resembles that of time domain MVDR. The main difference is the initial need to Fourier transform the data:
\begin{gather}
\x_\text{BS}[n] = \B\x[n], \qquad \B\in \mathbb{C}^{N_b,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
\end{gather}
$\B$ here is usually referred to as a Butler matrix. 
Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).

@@

When computing $\eR$ we we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. Combined these steps balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

These challenges are very closely linked. The Quadro 6000 architecture is of compute capability 2.0, which means there are 48\,kB of shared memory (L1 cache) and 128\,kB of registers per SM~\cite{Nvidia2012}. This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}. This will expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (Little's law). By dividing the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory\todo{introduce memory structure -shared for data shared between threads}, and 24 stored in registers. This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.

@@

@@

We have tested our GPU implementation of the MVDR beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar being attached to the HUGIN AUV~\cite{Hansen2009}. To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}). Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth. The image was processed with MVDR parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.

@@

Throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense. This should be apparent by studying Table \ref{tabbandwidth}. It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU. Looking at global memory, we note that it is only able to move 1 float for every 30 floating point operation that the GPU is able to execute. Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.

The solution is make sure that each thread primarily access the shared memory (L1 cache) and registers. Unfortunately, these are scarce resources and 
\begin{table}[b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{:}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{arithmetic}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
Arithmetic throughput & 1.03 Tflop/s & & \hspace{30pt} 1 & 30 \\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 & 30 \\
Shared memory & & 257 Gfloats/s & 1 & 4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 & 2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000; Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
\end{table}

>>>>

The beamspace MVDR formulation very closely resembles that of time domain MVDR. The main difference is the initial need to Fourier transform the data:
\begin{gather}
\x_\text{BS}[n] = \B\x[n], \qquad \B\in \mathbb{C}^{N_b,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
\end{gather}
The number of angles to compute is here defined by $N_b$, $\B$, usually referred to as a Butler matrix
Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).

@@

When computing $\eR$ we we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. This procedure balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes very little memory.

% These challenges are very closely linked. The Quadro 6000 architecture is of compute capability 2.0, which means there are 48\,kB of shared memory (L1 cache) and 128\,kB of registers per SM~\cite{Nvidia2012}. This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}. This will expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (Little's law). By dividing the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory\todo{introduce memory structure -shared for data shared between threads}, and 24 stored in registers. This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.

@@

% The solution is make sure that each thread primarily access the shared memory (L1 cache) and registers. Unfortunately, these are scarce resources and

@@

To display the imaging capabilities of the mentioned beamformers, we have processed experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar being attached to the HUGIN AUV~\cite{Hansen2009}. The studied object is the 1500 dwt oil tanker wreck Holmengraa, lying at 77\,m depth outside of Horten, Norway. HISAS1030 is a synthetic apterture sonar, but was operated in sidescan mode here. Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth. The image was processed with MVDR parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.

@@

\begin{table}[b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{:}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{arithmetic}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
Arithmetic throughput & 1.03 Tflop/s & & \hspace{30pt} 1 & 30 \\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 & 30 \\
Shared memory & & 257 Gfloats/s & 1 & 4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 & 2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
\end{table}

One thing seems clear to us after we have learned from this work is that it is very hard to reach the peak performance of a GPU, and more often

. In our this learned from this process is how that each GPU thread have very small amounts of A thing we took home from this work 
But no matter how Throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense. This should be apparent by studying Table \ref{tabbandwidth}. It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU. Looking at global memory, we note that it is only able to move 1 float for every 30 floating point operation that the GPU is able to execute. Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
